How Tree-like are the data?

### 2019-04-12 phylotreelike.rmd
syncing
git pushing

The way to think about "phylogenetic signal" is the amount of correlation we see each pair of species (distance?). 
The distribution plot above just shows the amount of change each species evolved compared to the ancestral node.
Thus, the null tree will have the same species level trait distribution, but what we are interested in are the "crosses".

Given a set of traits, we can now calculate the likelihood under the known/true tree (with the covariance) and the null tree (without the covariance).
We can do a likelihood ratio test and compute the Chisq statistic.

```{r likelihood surface, warning=FALSE, echo=FALSE, message=FALSE}

pp <- expand.grid(t1 = seq(-10,10,length=21)
  , t2 = seq(-10,10,length=21)    ## non-focal parameter
  , t3 = seq(-3,3,length=9)
)

@@

Stats 101, can we see it more clearly if we have more data?
There are two ways to increase data:

@@

pp <- expand.grid(t1 = seq(-10,10,length=9)
  , t2 = seq(-10,10,length=9)    ## non-focal parameter
  , t3 = seq(-10,10,length=9)
)

@@

Note: Maybe the extreme examples are not good?

```{r more species, warning=FALSE, echo=FALSE, message=FALSE,cache=TRUE}

ddprop <- data.frame(nspp=3:6, prop=NA)
ddprop$prop[1] <- mean(pp2$LRT)
rep_pp <- pp
for(i in 4:6){
phy <- rtree(n = i)
Vphy <- vcv(phy)

temp_dimension <- expand.grid(rep(1,nrow(rep_pp)),new=seq(-10,10,length=9))
rep_pp <- rbind(rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp,rep_pp)
newgrid <- cbind(rep_pp,temp_dimension[,"new"])
colnames(newgrid)[i] <- paste("t",i,sep="_")

newgrid$ll <- sapply(1:nrow(newgrid),function(x){mvtnorm::dmvnorm(newgrid[x,1:i], mean =rep(0,i), sigma = Vphy, log = TRUE)})
c <- qchisq(0.95,df=i-2)

newgrid$null <- sapply(1:nrow(newgrid),function(x){mvtnorm::dmvnorm(newgrid[x,1:i], mean =rep(0,i), sigma = Vnull, log = TRUE)})

newgrid2 <- (newgrid
  %>% rowwise()
  %>% mutate(LR = ll-null
             # , LRT = max(as.numeric(LR > log(c)),0.5)
             , LRT = 2*LR > log(c)
             )
  # %>% arrange(LR)
)

ddprop$prop[i-2] <- mean(newgrid2$LRT)
print(i)
rep_pp <- newgrid2[,1:i]
}

print(ddprop)
```

Brute forcing the grid is not a good idea and very inefficient. Can we get away with just sampling the tree?

@@

Y_phy <- MASS::mvrnorm(n=1000
  , mu=rep(0,each=i)
  , Sigma=Vphy
)

@@

Wow, it is surprising how fast it levels off.

@@

```{r increasing traits, echo=FALSE,cache=TRUE}

@@

I think I got a good idea now.

>>>>

The way to think about "phylogenetic signal" is the amount of correlation we see each pair of species (distance?). 
The distribution plot above just shows the amount of change each species evolve compared to the ancestral node.
Thus, the null tree will have the same species level trait distribution, but what we are interested in are the correlation (crosses).

Given a set of traits, we can now calculate the likelihood under the known/true tree (with the covariance) and the null tree (without the covariance).
We can do a likelihood ratio test and compute the Chisq statistic.
TODO: What is the "right" statistic? I am using Chisq for the easy examples and using Z when I am lazy.

```{r likelihood surface, warning=FALSE, echo=FALSE, message=FALSE, cache=TRUE}

pp <- expand.grid(t1 = seq(-10,10,length=51)
  , t2 = seq(-10,10,length=51)    ## non-focal parameter
  , t3 = seq(-3,3,length=9)
)

@@

Statistics 101, can we see it more clearly if we have more data?
There are two ways to increase data:

@@

pp <- expand.grid(t1 = seq(-10,10,length=25)
  , t2 = seq(-10,10,length=25)    ## non-focal parameter
  , t3 = seq(-10,10,length=25)
)

@@

Note: Maybe the extreme examples are not good? 
We still see a large area in the likelihood surface that aren't tree-like.

Brute forcing the grid is not a good idea and very inefficient when we increase the dimension. 
The grid itself is not a good representation of the sampling distribution anyway, thus when calculating the proportion of tree-like, it decreases as we increase the number of species.

Can we get away with just sampling traits from the tree?

@@

Y_phy <- MASS::mvrnorm(n=10000
  , mu=rep(0,each=i)
  , Sigma=Vphy
)

@@

Here I sample 10,000 points from the n-dimension space using a MVN with the phylogenetic variance. 
It is surprising how fast it levels off.

@@

```{r increasing traits, echo=FALSE,cache=TRUE,eval=FALSE}

@@

Here, I simulate 1000 sets of traits (lines) and did the LRT for each line. 
The first plot is consistent with the result before we see a small proportion of tree-like data.
In order to "increase" more data, I computated did the likelihood ratio test given we saw multiple sets of traits.
We can easily do this by adding the log likelihood for each set of traits under the NVM under the true tree and null tree and do the corresponding LRT. 
Note: Order does matter! We need a random order!

We can see that if we combine multiple traits, it increases our chance to detect if the data are tree-like via LRT.

Just a exteme example, what if we combine a bunch of non-tree-like traits? 
By ordering the
